The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Innate nature or tidal nurture?

### 2025-10-17 thesis/acknowledgments.tex

To Raphaël, who has helped inspire and direct this project. To Isabel
and Ryan.

To Jax, who has been a support along the way, guiding my research
methods and project. Akshara and Ewould.

To all of my new friends I have made along the way. Ruxin, Aviv, Anya,
Pranav, Ted, Adrianne, Will, Maeve, and everyone else I have met. We
have build a wonderful community of students in this department. And to
my friends across UVic, Wyatt, Rosemary, Maggie,

To my past advisors, James who has continued to support me long after
our paths diverged, David and Jennifer for guiding my first research
project.

To Wayne for showing us the joy of astronomy and outreach. To my mentor,
Ebtihal.

@@

To Fierce Brosnen and our trivia team. Melody, Lily, Gillian, Stephen,
Jillian, and so many more.

@@

To my musical friends, Adrienne, Alex for our shared joy in music
making. And Steve, for showing me the joy in a simple life and beautiful
music. To my fellow knitters. Anne, Anya, Leise, Wes. To the
international community.

To my parents for their continuous support throghout this project. To my
grandparents for always being there. To my cousins and my extended
family. And to my siblings for always being a source of joy.

To the stars and the moon. To the land of Victoria, and its historical
caretakers. To the waters and beautiful trees, the ocean waves, rocks
and hills and jagged coastlines. To the rich, enveloping forests and
rust-red arbutus tress, scraggly Garry oaks and Camus flowers. I could
not imagine a more beautiful place to study and to always remind my
place in the world.

>>>>

To Raphaël, who has helped inspire and direct this project. To everyone
who has helped provide me opportunities to share work, and share
wonderful ideas with me. Isabel, Jens, Eduardo, Jorge, David, Justin,
Eugene, Azi, José, and many others.

To Jax, who has been a support along the way, guiding my research
methods and project.

To all of my new friends I have made along the way. Ruxin, Aviv. To
Pranav, Ted, Jono, Adrianne, Praneet, Will, Maeve, Ewoud. To Akshara for
all your personal and professional advice. We have build a wonderful
community of students in this department. To all the students who have
since moved on and show what we may be: Bobby, Leo, Ryan, Katie,
Fletcher, Ayden, Maheyer, Jess, Dhwani, Spencer, Mallory, Omar, Isaac,
and Dasha.

And to my friends across UVic, Wyatt, Rosemary, Maggie, and the many
people I am still meeting.

To the Linguistics extended family and team \emph{adequate}. Melody,
Lily, Gillian, Stephen, Jamie, Chloë, Alyssa, Kyle, and Marcus. You are
all wonderful human beings, cooks, and always a joy to spend time with.

To my musical friends, Adrienne, Alex for our shared joy in music
making. And Steve, for showing me the joy in a simple life and beautiful
music.

To the Victoria performing arts community and Fierce Brosnan, our local
legend.

To my fellow crafters, knitters, and crochet friends. Leise, Wes, Anne,
Anya, Ashley, Sabrina, Sarah. To the graduate student society for
continuing to connect us across departments.

To the international community, and my international friends---Marcela,
Bernardo, Daniel, Walla, Mehrab, and everyone else I have met.

To Erica. To my students and everyone working with outreach here. I love
being able to pass on our knowledge to the wider world.

To my TEP friends---I hope you all continue to do well.

To all of my grade school math nerds---Stanley, Alex, Rick, Andrew. I
enjoyed all of our adventures together, appreciate helping myself bloom
in curiosity and exploration, and keep on math-ing!

To my past advisors, James who has continued to support me long after
our paths diverged, David and Jennifer for guiding my first research
project. To Wayne for showing us the joy of astronomy and outreach. To
my mentor, Ebtihal.

@@

@@

To my parents for their continuous support throughout this project. To
my grandparents for always being there.

To all of my cousins, but especially Bridget and Gretchen, and my
extended family. And to my siblings for always being a source of joy. To
Arya. To all of my furry friends I have sat---Nano, Pluto, Hobbs,
Penelope, Tochka, Camar, Loki, and Hermes.

To the stars and the moon, for your enduring beauty, continuing to
inspire me and posing unending questions about our natural world.

To the land of Victoria, and its historical caretakers. To the waters
and beautiful trees, the ocean waves, rocks and hills and jagged
coastlines. To the rich, enveloping forests and rust-red arbutus tress,
scraggly Garry oaks and Camus flowers. I could not imagine a more
beautiful place to study and to always remind my place in the world.

### 2025-10-17 thesis/data.tex

As discussed in Section~\ref{sec:exponential_profiles}, the projected
luminosity/stellar mass density profile of dwarf galaxies is generally
well-described by an exponential law (Eq.~\ref{eq:exponential_law}). A
prototypical example is the Fornax dSph --- well-fit by an exponential
over 4.5 decades in surface density. On the other hand, Sculptor or Ursa
Minor have profiles that deviate significantly from an exponential
profile fitted to the inner regions, with a clear excess of stars/light
in the outer regions. In this chapter, we critically review the density
profile of the MW classical dwarfs. This thesis concerns the origin of
the outer deviations from exponential profiles, with a tidal
interpretation explored in subsequent chapters.

@@

Our density profiles are robust to alternative methodologies and
magnitude cuts. Density profiles based on J+24 candidates may be biased
by their assumed spatial likelihoods. To address this, we consider a
Bayesian model with a non-parametric spatial likelihood. We also explore
selections using absolute cuts, and the effect of structural assumptions
of the density profile in Appendix \ref{sec:extra_density}. These
methodologies are consistent except when the derived density drops below
the background of satellite-like stars (from the CMD+PM sample). We also
do not find evidence of magnitude biases in the density profiles,
indicating that inhomogeneous magnitude-completeness is small. Our
density profiles furthermore agree with photometric surveys and similar
literature (see Appendix \ref{sec:extra_density}). We conclude that
selection criteria do not influence our results.

>>>>

As discussed in Section~\ref{sec:exponential_profiles}, the projected
luminosity/stellar mass density profile of dwarf galaxies is generally
well-described by an exponential law (Eq.~\ref{eq:exponential_law}). A
prototypical example is the Fornax dSph --- well-fit by an exponential
over 4.5 decades in surface density. On the other hand, Sculptor or Ursa
Minor have profiles that deviate significantly from an exponential
profile fitted to the inner regions, with a clear excess of stars/light
in the outer regions. In this Chapter, we critically review the density
profile of the MW classical dwarfs. This thesis concerns the origin of
the outer deviations from exponential profiles, with a tidal
interpretation explored in subsequent Chapters.

@@

Our density profiles are robust to alternative methodologies and
magnitude cuts. Density profiles based on J+24 candidates may be biased
by their assumed spatial likelihoods. To address this, we consider a
Bayesian model with a non-parametric spatial likelihood. We also explore
selections using absolute cuts, and the effect of structural assumptions
of the density profile in Appendix \ref{sec:extra_density}. These
methodologies are consistent except when the derived density drops below
the background of satellite-like stars (from the CMD+PM sample). We also
do not find evidence of magnitude biases in the density profiles,
indicating that inhomogeneous magnitude-completeness is small. Our
density profiles furthermore agree with photometric surveys and similar
literature (see Appendix \ref{sec:extra_density}). We conclude that
selection criteria do not influence our conclusions.

### 2025-10-17 thesis/discussion.tex

In the previous chapters, we have shown that Sculptor and Ursa Minor
have more extended density profiles compared to other MW classical
satellites, and that recent tides are an unlikely explanation for their
extended outer structure. We therefore suggest that Scl and UMi may host
an additional, extended population of stars, such as a ``stellar halo''.
With our data, other classical dwarfs lack evidence for similar extended
populations. Thus, Scl and UMi have followed a divergent evolutionary
path, motivating our discussion of extended density formation pathways
here.

In this chapter, we place our results in context with past work. We
review both works, considering tidal effects on Scl and UMi and
detections of peculiarities. We then consider possible origins of
extended stellar populations for these galaxies. Scenarios include
multi-epoch star formation, past mergers depositing stars into the
outskirts, and even departures from \LCDM{}. We later discuss how future
observations and simulations may distinguish among formation scenarios.
Finally, we summarize the main conclusions of this thesis.

@@

Tidal effects have long been invoked to explain features of dwarf
galaxies \citep[e.g.,
Section~\ref{sec:scl_umi_obs_tides},][]{mayer+2001a, tsujimoto+shigeyama2002}.
Yet, growing work suggests tides may not be as commonplace as suspected.
For example, \citet{read+2006} suggested that the absence of rising
radial velocity dispersion profiles in dwarfs implies a lack of tidal
disruption. \citet{penarrubia+2009}, introducing the break radius as a
diagnostic (Eq.~\ref{eq:r_break}), showed that most satellite orbits
were inconsistent with observable tidal effects. Recently,
\citet{pace+erkal+li2022} used a Jacobi radius-based criterion
(Eq.~\ref{eq:r_jacobi}) to reach a similar conclusion. Except for the
Sagittarius dSph, these results suggest tides have a mild influence on
the stellar components of classical dwarfs.

A few works have modelled Scl or UMi specifically, finding similar
conclusions. \citet{iorio+2019} applied idealized N-body simulations to
study tidal effects on Scl. They similarly found weak tidal effects,
even for a dark-matter-free model. Most recently,
\citet{tchiorniy+genina2025} also used idealized simulations tuned to
five classical dwarfs with a focus on the inner density, concluding that
tides do not strongly affect equilibrium assumptions or the stellar
component.

@@

Besides extended stellar density profiles, Scl and UMi display other
peculiarities that may hold clues to their formation. Scl and UMi both
host at least two chemodynamic populations, as revealed through their
photometric or metallicity-velocity structure\footnote{Other examples of
  galaxies with multiple populations include Carina
  \citep{battaglia+2012, fabrizio+2016, kordopatis+2016}, Fornax
  \citep{battaglia+2006, amorisco+evans2012, delpino+aparicio+hidalgo2015},
  Sextans
  \citep{battaglia+2011, cicuendez+battaglia2018, roederer+2023}, and
  Andromeda II
  \citep{mcconnachie+arimoto+irwin2007, ho+2012, delpino+2017}.}
\citep{tolstoy+2004, battaglia+2008, pace+2020}. The inner population is
younger, higher metallicity, and dynamically colder, whereas the outer
population is older, lower metallicity, and dynamically hotter.

Could these internal stellar populations map to the observed density
excess? In Scl, both populations have scale radii likely smaller than
the half-light radius. But, the third, more tentatively detected
population in \citet{arroyo-polonio+2024} would reside within the outer
density excess. In Ursa Minor, the transition between the metal-rich and
metal-poor stellar components occurs at \(\sim 30\,\)arcmin
\citep{pace+2020}, which coincides with the density excess. This
metal-poor population may thus be related to the extended stellar
density of Ursa Minor.

@@

\textbf{Major mergers.} Dwarf galaxy mergers may be relatively common
\citep{deason+wetzel+garrison-kimmel2014}. After a merger, stars from
the lower-mass galaxy are preferentially dispersed, forming an extended
stellar component and population gradient
\citep{benitez-llambay+2016, deason+2022}. If the galaxies contain gas,
the merger can trigger new star formation, forming a younger population
of stars \citep[e.g.,][]{genina+2019}. A few local dwarfs are suspected
to have undergone a major merger, including Tucana II, Andromeda II, and
Phoenix
\citep{lokas+2014, fouquet+2017, tarumi+yoshida+frebel2021, cardona-barrero+2021, querci+2025}.
A galaxy having undergone a major merger should harbour at least two or
three populations from distinct origins.

@@

In both cases, tides more plausibly produce extended density profiles.
However, recovering the observed velocity dispersions often requires
ongoing tidal disruption
\citetext{\citealp{mcgaugh+wolf2010}; \citealp[but see
also][]{sanchez-salcedo+hernandez2007}}. Current data do not show
evidence of such features (see Appendix
\ref{sec:extra_rv_models})---future data would need to uncover signs of
disruption to support this hypothesis.

\textbf{Dynamical heating.} Old stars in dwarf galaxies may be hotter
than younger stars due to processes including stellar feedback
\citep{stinson+2009, maxwell+2012, el-badry+2016, mercado+2021},
sub-subhalo interactions \citep{penarrubia+2025}, or even fuzzy dark
matter interference fringes
\citep[e.g.,][]{el-zant+2020, duttachowdhury+2023}. However, most of
these processes should operate similarly across dwarf
galaxies---extended stellar populations should be more common if these
processes are important.

@@

We reviewed a number of different, possibly concurrent explanations for
the formation of extended populations. While we leave the nature of Scl
and UMi's extended stellar densities an open question, we can discuss
possible clues to different formation scenarios.

@@

In this thesis, we have investigated the extended, outer density excess
of Sculptor and Ursa Minor and its possible tidal origin. We show that
the density profiles are robust, implying that Scl and UMi contain a
true density excess relative to other classical dwarfs.

>>>>

In the previous chapters, we have shown that Sculptor and Ursa Minor
have more extended density profiles compared to other MW classical
satellites, and that recent tides are an unlikely explanation for their
extended outer structure. We therefore suggest the reason that Scl and
UMi have extended light profiles is that they host an additional,
extended population of stars, such as a ``stellar halo.'' With our data,
other classical dwarfs lack evidence for similar extended populations.
Thus, Scl and UMi have followed a divergent evolutionary path,
motivating our discussion of extended density formation pathways here.

In this chapter, we place our results in context with past work. We
review that work, considering tidal effects on Scl and UMi and
detections of peculiarities. We then consider possible origins of
extended stellar populations in these galaxies. Scenarios include
multi-epoch star formation, past mergers depositing stars into the
outskirts, and even departures from \LCDM{}. We later discuss how future
observations and simulations may distinguish among formation scenarios.
Finally, we summarize the main conclusions of this thesis.

@@

Tidal effects have long been invoked to explain features of dwarf
galaxies \citep[e.g.,
Section~\ref{sec:scl_umi_obs_tides},][]{mayer+2001a, tsujimoto+shigeyama2002}.
Yet, a growing body of work suggests tides may not be as commonplace as
suspected. For example, \citet{read+2006} suggested that the absence of
rising radial velocity dispersion profiles in dwarfs implies a lack of
tidal disruption. \citet{penarrubia+2009}, introducing the break radius
as a diagnostic (Eq.~\ref{eq:r_break}), showed that most satellite
orbits were inconsistent with observable tidal effects. Recently,
\citet{pace+erkal+li2022} used a Jacobi radius-based criterion
(Eq.~\ref{eq:r_jacobi}) to reach a similar conclusion. Except for the
Sagittarius dSph, these results suggest tides have a mild influence on
the stellar components of classical dwarfs.

A few studies have modelled Scl or UMi specifically, finding similar
conclusions. \citet{iorio+2019} applied idealized N-body simulations to
study tidal effects on Scl. They similarly found weak tidal effects,
even for a dark-matter-free model. Most recently,
\citet{tchiorniy+genina2025} also used idealized simulations tuned to
five classical dwarfs with a focus on the inner density, concluding that
tides do not strongly affect equilibrium assumptions or the stellar
component.

@@

Besides extended stellar density profiles, Scl and UMi display other
peculiarities that may hold clues to their formation. Scl and UMi both
host at least two distinct chemodynamic populations, as revealed through
their photometric or metallicity-velocity structure\footnote{Other
  examples of galaxies with multiple populations include Carina
  \citep{battaglia+2012, fabrizio+2016, kordopatis+2016}, Fornax
  \citep{battaglia+2006, amorisco+evans2012, delpino+aparicio+hidalgo2015},
  Sextans
  \citep{battaglia+2011, cicuendez+battaglia2018, roederer+2023}, and
  Andromeda II
  \citep{mcconnachie+arimoto+irwin2007, ho+2012, delpino+2017}.}
\citep{tolstoy+2004, battaglia+2008, pace+2020}. The inner population is
younger, higher metallicity, and dynamically colder, whereas the outer
population is older, lower metallicity, and dynamically hotter.

Could these internal stellar populations map to the observed density
excess? In Scl, both populations have scale radii likely smaller than
the half-light radius. But, a third, more tentatively detected
population in \citet{arroyo-polonio+2024} would reside within the outer
density excess. In Ursa Minor, the transition between the metal-rich and
metal-poor stellar components occurs at \(\sim 30\,\)arcmin
\citep{pace+2020}, which coincides with the density excess. This
metal-poor population may thus be related to the extended stellar
density of Ursa Minor.

@@

\textbf{Major mergers.} Dwarf galaxy mergers may be relatively common
\citep{deason+wetzel+garrison-kimmel2014}. After a merger, stars from
the lower-mass galaxy are preferentially dispersed, forming an extended
stellar component and population gradient
\citep{benitez-llambay+2016, deason+2022}. If the galaxies contain gas,
the merger can trigger new star formation, forming a younger population
of stars \citep[e.g.,][]{genina+2019}. A few local dwarfs are suspected
to have undergone a major merger, including Tucana II, Andromeda II, and
Phoenix
\citep{lokas+2014, fouquet+2017, tarumi+yoshida+frebel2021, cardona-barrero+2021, querci+2025}.
A galaxy having undergone a major merger should harbour at least two or
three populations with distinct origin.

@@

In both cases, tides more plausibly produce extended density profiles.
However, recovering the observed velocity dispersions often requires
ongoing tidal disruption
\citetext{\citealp{mcgaugh+wolf2010}; \citealp[but see
also][]{sanchez-salcedo+hernandez2007}}. Current data do not show
evidence of such features (see Appendix
\ref{sec:extra_rv_models})---future data would be needed to uncover
signs of disruption to support this hypothesis.

\textbf{Dynamical heating.} Old stars in dwarf galaxies may be
dynamically hotter than younger stars due to processes including stellar
feedback
\citep{stinson+2009, maxwell+2012, el-badry+2016, mercado+2021},
sub-subhalo interactions \citep{penarrubia+2025}, or even fuzzy dark
matter interference fringes
\citep[e.g.,][]{el-zant+2020, duttachowdhury+2023}. However, most of
these processes should operate similarly across dwarf
galaxies---extended stellar populations should be more common if these
processes are important.

@@

We have reviewed above a number of different, possibly concurrent
explanations for the formation of extended populations. While we leave
the nature of Scl and UMi's extended stellar densities an open question,
we can discuss possible clues to different formation scenarios.

@@

In this thesis, we have investigated the extended, outer density excess
of Sculptor and Ursa Minor and its possible tidal origin. We have shown
that the density profiles are robust, implying that Scl and UMi contain
a true density excess relative to other classical dwarfs.
